---
title: "Analysis of Social Media Usage Among Students"
author: "Tina Mariyam Jacab"
enrollment_no: "2505101270264"
# Enrollment No: 2505101270264
date: "2026-09-23"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

# 1. Introduction

Social media is commonly used by students for communication, entertainment, education and socializing. This Tiny Project analyzes a small fictional student dataset to understand daily social media usage and its relationship with study time and academic performance.

## Objective

The main objectives are:

- Import student social media data from a CSV file.
- Check and clean missing values.
- Calculate basic descriptive statistics.
- Compare usage across platforms and purposes.
- Create graphs and plots.
- Examine the relationship between social media usage and academic performance.
- Interpret the results and present findings.

# 2. Data Importing

The dataset contains 30 fictional students and includes gender, primary social media platform, daily social media hours, study hours, sleep hours, main purpose and academic performance.

```{r import-data}
student_data <- read.csv("social_media_usage_students.csv")
head(student_data)
str(student_data)
summary(student_data)
```

**Code explanation:** `read.csv()` imports the CSV file. `head()` displays the first rows, `str()` shows the structure and data types, and `summary()` gives a basic statistical summary.

# 3. Data Preprocessing and Cleaning

```{r cleaning}
# Check missing values
colSums(is.na(student_data))

# Replace missing numeric values with the column median
student_data$Daily_Social_Media_Hours[
  is.na(student_data$Daily_Social_Media_Hours)
] <- median(student_data$Daily_Social_Media_Hours, na.rm = TRUE)

student_data$Sleep_Hours[
  is.na(student_data$Sleep_Hours)
] <- median(student_data$Sleep_Hours, na.rm = TRUE)

# Check again
colSums(is.na(student_data))
```

**Code explanation:** `is.na()` identifies missing values. The missing numeric observations are replaced with their respective median values. `na.rm = TRUE` tells R to ignore missing values while calculating the median.

# 4. Statistical Analysis

```{r statistics}
mean_social <- mean(student_data$Daily_Social_Media_Hours)
median_social <- median(student_data$Daily_Social_Media_Hours)
sd_social <- sd(student_data$Daily_Social_Media_Hours)

mean_study <- mean(student_data$Daily_Study_Hours)
mean_marks <- mean(student_data$Academic_Performance)

mean_social
median_social
sd_social
mean_study
mean_marks
```

### Main results

- Mean social media usage: **3.10 hours/day**
- Median social media usage: **3.00 hours/day**
- Standard deviation: **1.03 hours**
- Mean study time: **2.55 hours/day**
- Mean academic performance: **77.00**

**Interpretation:** The average student in this sample spends about 3.10 hours per day on social media. The standard deviation shows how much individual usage differs from the average.

# 5. Platform Analysis

```{r platform}
platform_count <- table(student_data$Primary_Platform)
platform_count

platform_avg <- aggregate(
  Daily_Social_Media_Hours ~ Primary_Platform,
  data = student_data,
  FUN = mean
)
platform_avg

barplot(
  platform_avg$Daily_Social_Media_Hours,
  names.arg = platform_avg$Primary_Platform,
  main = "Average Daily Social Media Usage by Platform",
  xlab = "Primary Platform",
  ylab = "Average Hours per Day"
)
```

**Interpretation:** The bar chart compares average daily usage among the main platforms represented in the dataset.

# 6. Distribution of Social Media Usage

```{r histogram}
hist(
  student_data$Daily_Social_Media_Hours,
  main = "Distribution of Daily Social Media Usage",
  xlab = "Hours per Day",
  ylab = "Number of Students"
)
```

**Interpretation:** The histogram shows how students are distributed across different levels of daily social media usage.

# 7. Purpose of Social Media Use

```{r purpose}
purpose_count <- table(student_data$Main_Purpose)
purpose_count

pie(
  purpose_count,
  main = "Main Purpose of Social Media Use",
  labels = paste(names(purpose_count), purpose_count)
)
```

**Interpretation:** The pie chart shows the major purposes for which students report using social media.

# 8. Relationship with Academic Performance

```{r correlation}
correlation <- cor(
  student_data$Daily_Social_Media_Hours,
  student_data$Academic_Performance
)

correlation

plot(
  student_data$Daily_Social_Media_Hours,
  student_data$Academic_Performance,
  main = "Social Media Usage vs Academic Performance",
  xlab = "Social Media Hours per Day",
  ylab = "Academic Performance",
  pch = 19
)
```

The calculated correlation for this dataset is approximately **-0.962**.

**Interpretation:** The correlation describes the direction and strength of the linear relationship in this sample. A correlation value by itself does not prove that social media usage causes changes in academic performance.

# 9. Relationship with Study Time

```{r study-correlation}
cor(
  student_data$Daily_Social_Media_Hours,
  student_data$Daily_Study_Hours
)
```

For this sample, the correlation between social media hours and study hours is approximately **-0.940**.

# 10. Findings

1. Students in the sample use social media for several purposes, including communication, education, entertainment and socializing.
2. Average daily social media usage is approximately **3.10 hours**.
3. Social media usage varies between students, as shown by the standard deviation of **1.03 hours**.
4. The platform comparison shows differences in average usage.
5. The correlation analysis describes the association between social media usage, study time and academic performance in this sample.

# 11. Conclusion

This Tiny Project demonstrates how R can be used to import, clean, analyze and visualize student social media usage data. The analysis provides a simple statistical overview of usage patterns and their relationships with study time and academic performance.

**Note:** The dataset is fictional and is intended only for academic demonstration. The results should not be generalized to all students.

# 12. R Code Summary

The project uses basic R functions such as `read.csv()`, `is.na()`, `mean()`, `median()`, `sd()`, `aggregate()`, `table()`, `barplot()`, `hist()`, `pie()`, `plot()` and `cor()` to complete the analysis.
